---
title: "Statistical Modeling of Medical Charges and Health Risk"
author: "Jiyu Zhang"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# 📌 Introduction

This project investigates the relationship between patient characteristics (e.g., age, BMI, smoking status) and healthcare cost using regression analysis, as well as the probability of health risk using generalized linear models (GLMs).

Two tasks are conducted:
1. Linear regression modeling of medical charges.
2. Logistic regression modeling of binary health risk.

The data used here is simulated for demonstration purposes and shares a similar structure with the original coursework dataset.

---

# 📊 Task 1 – Linear Regression Analysis of Medical Charges
```{r}
setwd("/Users/jiyuzhang/Desktop/Heriot-Watt/F79MB/Project\ 2")
data1=read.csv("charges.csv",header=T)
attach(data1)
###################################1#############################################################
#(a)judge the distribution approximately
smokers=numeric(length(smoker))
for (i in 1:length(smoker)) {
  if(smoker[i]!="yes"){
    smokers[i]=0
  }else if (smoker[i]!="no"){
    smokers[i]=1
  }
}
smokers=as.numeric(as.character(smokers))
par(mfrow=c(2,2))
#plot the histograms
hist(age,col="red",breaks = 10)#uniform distribution
hist(bmi,col="red",breaks = 20)#normal distribution
hist(smokers,col="red",breaks = 10)#binomial distribution
hist(charges,col="red",breaks = 10)#exponential distribution
###############
#Q-Q plot to fit the distribution
mean_age=mean(age)
mean_bmi=mean(bmi)
mean_smokers=mean(smokers)
mean_charges=mean(charges)
sd_age=sd(age)
sd_bmi=sd(bmi)
sd_smokers=sd(smokers)
sd_charges=sd(charges)
#for age
par(mfrow=c(2,2))
plot(qunif(seq(from=0,to=1,by=0.01),min=min(age),max=max(age)),xlab="U(40.12521)",ylab="data of age",quantile(age,probs = seq(from=0,to=1,by=0.01)),col="red",lwd="1",pch=19,main="Q-Q plot for age")
abline(0,1,col="black",lwd="2")
#for bmi
plot(qnorm(seq(from=0,to=1,by=0.01),mean=mean_bmi,sd=sd_bmi),quantile(bmi,probs = seq(from=0,to=1,by=0.01)),col="red",lwd="1",pch=19,xlab="N(31.12883,6.118566^2)",ylab="data of bmi",main="Q-Q plot for bmi")
abline(0,1,col="black",lwd="2")
#for smoker
plot(qbinom(seq(from=0,to=1,by=0.01),size=1,prob=mean_smokers),quantile(smokers,probs = seq(from=0,to=1,by=0.01)),xlab="Bin(0.1783877)",ylab="data of smoker",col="red",lwd="1",pch=19,main="Q-Q plot for smoker")
abline(0,1,col="black",lwd="2")
#for charges
plot(qexp(seq(from=0,to=1,by=0.01),rate=1/mean_charges),quantile(charges,probs = seq(from=0,to=1,by=0.01)),xlab="Exp(1/12923.02)",ylab="data of charges",col="red",lwd="1",pch=19,main="Q-Q plot for charges")
abline(0,1,col="black",lwd="2")

#(b)
#Pairs the data to plot
bbb=data.frame(charges,age,bmi)
pairs(bbb,col="red",lwd="0.5",pch=19,main="pairs of charges, age and bmi")

#(c)
par(mfrow=c(2,2))
#fit the model with all explanatory variables with interpretation terms
model1=lm(charges~(age*smoker+bmi*smoker))
summary(model1)
plot(model1,which=1:4)
summary.aov(model1)
#fit model removing insignificant terms
model2=lm(charges~age+bmi*smoker)#best
summary(model2)
plot(model2,which=1:4)
summary.aov(model2)
#(d)
JJ=predict(model2, data.frame(age=44,bmi=22.5,smoker="yes"),interval = "confidence",level=0.95,type="response");JJ#convidence interval
#Task2
#load the data of task 2
setwd("/Users/jiyuzhang/Desktop/Heriot-Watt/F79MB/Project\ 2")
data2=read.table("CHDRisk.txt", header=T)
attach(data2)
Genders=numeric(length(Gender))
for (i in 1:length(Gender)) {
  if(Gender[i]!="M"){
    Genders[i]=0
  }else if (Gender[i]!="F"){
    Genders[i]=1
  }
}
Genders=as.numeric(as.character(Genders))
#(a)
par(mfrow=c(2,2))
boxplot(Genders[Risk==0],Genders[Risk==1],col="red",xlab="Risk=0                   Risk=1",ylab="Gender")
boxplot(Age[Risk==0],Age[Risk==1],col="red",xlab="Risk=0                   Risk=1",ylab="Age")
boxplot(Totchol[Risk==0],Totchol[Risk==1],col="red",xlab="Risk=0                   Risk=1",ylab="Totchol")
boxplot(Glucose[Risk==0],Glucose[Risk==1],col="red",xlab="Risk=0                   Risk=1",ylab="Glucose")

#(b)
u1=0
u2=0
u3=0
u4=0
for (i in 1:length(Risk)) {
  if(Gender[i]!="F"&&Risk[i]>0){
    u1=u1+1
  }
}
for (i in 1:length(Risk)) {
  if(Gender[i]!="F"&&Risk[i]<1){
    u2=u2+1
  }
}
for (i in 1:length(Risk)) {
  if(Gender[i]!="M"&&Risk[i]>0){
    u3=u3+1
  }
}
for (i in 1:length(Risk)) {
  if(Gender[i]!="M"&&Risk[i]<1){
    u4=u4+1
  }
}
U=matrix(c(u1,u2,u3,u4),nrow=2)#observation quantities
P=matrix(c(141*52/317,141*265/317,176*52/317,176*265/317),nrow=2)#expected quantities
chisq.test(U, p = P,correct=FALSE)#without Yate's correction
chisq.test(U, p = P,correct=TRUE)#with Yate's correction

#(c)
#fit the model with all explanatory variables
par(mfrow=c(2,2))
model9=glm(Risk~Age+Totchol+Glucose+Gender,family=binomial)#best
summary(model9)
plot(model9,which=1:4)
anova(model9)
aov(model9)
#fit model by removing insignificant terms
model10=glm(Risk~Age+Glucose+Gender,family=binomial)
summary(model10)
plot(model10,which=1:4)

model11=glm(Risk~Age+Totchol+Gender,family=binomial)
summary(model11)
plot(model11,which=1:4)

model12=glm(Risk~Age+Gender,family=binomial)
summary(model12)
plot(model12,which=1:4)
#use chi-square test to compare the model with the maximal model
chi_5=1 - pchisq(237.82, df=312);chi_5
chi_6=1 - pchisq(237.88, df=313);chi_6
chi_7=1 - pchisq(238.63, df=313);chi_7
chi_8=1 - pchisq(238.78, df=314);chi_8
#use chi-square test to compare the model with the first model with all explanatory variables
chi_9=1 - pchisq(237.82-237.82, df=312-312);chi_9
chi_10=1 - pchisq(237.88-237.82, df=313-312);chi_10
chi_11=1 - pchisq(238.63-237.82, df=313-312);chi_11
chi_12=1 - pchisq(238.78-237.82, df=314-312);chi_12
chi_13=1 - pchisq(282.96-238.78, df=316-314);chi_13
#plot the deviance residuals plot for the model12
par(mfrow=c(1,2))
plot(residuals(model12,type="deviance")[Gender=="M"] ~ Age[Gender=="M"],xlab="Age",ylab="Deviance residuals")
points(residuals(model12,type="deviance")[Gender=="F"] ~ Age[Gender=="F"],col="red")
legend(x="topright",legend=c("Male","Female"),pch=c(21,21),lty=c(1,1),col=c("black","red"))
plot(residuals(model12,type="deviance")[Gender=="M"] ~ fitted.values(model12,type="deviance")[Gender=="M"],xlab="Fitted values",ylab="Deviance residuals")
points(residuals(model12,type="deviance")[Gender=="F"] ~ fitted.values(model12,type="deviance")[Gender=="F"],col="red")
legend(x="topright",legend=c("Male","Female"),pch=c(21,21),lty=c(1,1),col=c("black","red"))

#d
KK=predict(model12, data.frame(Age=83,Totchol=105,Glucose=218,Gender="M"), type="response");KK


```